A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
impactos

@@

```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
q <- quantile(provinces_spain$Density)
provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
p1 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$Older)
provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
p2 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$GDPpc)
provinces_spain$Quantile<- as.factor((provinces_spain$GDPpc > q[2]) + (provinces_spain$GDPpc > q[3]) +(provinces_spain$GDPpc >= q[4]) + 1)
p3 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
gridExtra::grid.arrange(p1,p2,p3,nrow=2)
```

@@

@@

>>>>

<!-- load cache -->
<!-- whoppeR::load_knitr_cache('/Users/fernandoair/Dropbox/COVID-19/covid19-environmental-correlates/Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain_cache/latex/') -->

@@

```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
q <- quantile(provinces_spain$Density)
provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
p1 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$Older)
provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
p2 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$GDPpc)
provinces_spain$Quantile<- as.factor((provinces_spain$GDPpc > q[2]) + (provinces_spain$GDPpc > q[3]) +(provinces_spain$GDPpc >= q[4]) + 1)
p3 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))

provinces_spain$M.Transit<- as.factor(provinces_spain$Transit)
p4 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = M.Transit),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("white", "gray1"))
gridExtra::grid.arrange(p1,p2,p3,p4,nrow=2)
```

@@

```{r impacts}
Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
trsur <- spatialreg::trW(Wsur, type = "MC")
impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)

impacts.sur.slm_lag11[[1]]
impacts.sur.slm_lag11[[2]]
impacts.sur.slm_lag11[[3]]

myimpact <- matrix(0,ncol = 5, nrow = 29)
for (i in 2:30){
  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
}
matplot(myimpact, type='l')

@@

* HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA DARLE RELEVANCIA A LA ESTIMACION SUR

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
impactos

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatial analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

Two set of variables has been selected to explain the diary incidence
ratio in each Spanish province. The first, are climatic variables
(temperature, humidity and sunshine) and its impact has been argued in
the introduction section. The second, is a set of specific
characteristics of each province. The objective is include in our model
some control variables (socioeconomic characteristics) to capture the
probable heterogeneity of each spatial unit. All those variables are
constant along the temporal period analized in this research. In this
subsetion we present some evidences of the expected relationship between
the four control variables considered an the incidence ratio.

The structure of population is a key variable on the incidence. Several
papers related the relationship between a fast trasnmision of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC) and therefore a positive sign will be expected. Note that our
analysis is at provincial level - not a urban level - and therefore is
necessary to rethinking the relationship. In Spain the population is
concentrate (with som exception) in the mediterranean coast (Goerlich
and Cantarino (2013)), in general with high temperature and high
humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related
with the population structure include like control variable. Unlike the
population density the older people in Spain is concentrate in the core
of the Iberian Peninsula (see figure \ref{plotfer1}).

The per capita GDP also can generate differences between provinces. No
direct relationship can be observed between the GDP and the ratio of
incidence but the GDP is a proxy of inobservable information. This
varaible has been included in similar papers (Wang et al. (2020)). For
low DGPpc is more difficult maintain social distance (sing +); High
DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
difícl mantener el confinamiento.

@@

@@

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatial analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

Two set of variables has been selected to explain the daily incidence
ratio in each Spanish province. The first, are climatic variables
(temperature, humidity and sunshine) and its impact has been argued in
the introduction section. The second, is a set of specific
characteristics of each province. The objective is include in our model
some control variables (socioeconomic characteristics) to capture the
probable heterogeneity of each spatial unit. All those variables are
time constant along the temporal period analized in this research but
have high spatial variability. In this subsetion we present some
evidences of the expected relationship between the four control
variables and the incidence ratio.

The structure of population is a key variable on the incidence. Several
papers related the relationship between a fast trasnmision of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC) and therefore a positive sign will be expected. Note that our
analysis is at provincial level - not a urban level - and therefore is
necessary to rethinking the relationship. In Spain the population is
concentrate (with some exception) in the mediterranean coast (Goerlich
and Cantarino (2013)), in general with high temperature and high
humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related
with the population structure include like control variable. Unlike the
population density the older people in Spain is concentrate in the core
of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is
that in the provinces with more older adults the social contact is lower
(Prem et al. (2017)) and therefore the a negative correlation will be
expected. This hypotesis is based in the idea that the agee-structured
social contact matrix in Spain is similar to Italy (see Singh and
Adhikari (2020) Fig 1 pag 3)

The per capita GDP (GDPpc) also can generate differences between
provinces. No direct relationship can be observed between the GDP and
the ratio of incidence but our hypothesis is that the GDPpc is a proxy
of inobservable information. By example, provinces with high DGPpc have
more transit of foreing people This variable has been included in
similar papers (Wang et al. (2020)). For low DGPpc is more difficult
maintain social distance (sing +); High DGPpc -\textgreater{} more
economic activity -\textgreater{} es mas difícl mantener el
confinamiento.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Wsur <-}\StringTok{ }\KeywordTok{as}\NormalTok{(spdep}\OperatorTok{::}\KeywordTok{listw2mat}\NormalTok{(listw), }\StringTok{"CsparseMatrix"}\NormalTok{)}
\NormalTok{trsur <-}\StringTok{ }\NormalTok{spatialreg}\OperatorTok{::}\KeywordTok{trW}\NormalTok{(Wsur, }\DataTypeTok{type =} \StringTok{"MC"}\NormalTok{)}
\NormalTok{impacts.sur.slm_lag11 <-}\StringTok{ }\KeywordTok{impactspsur}\NormalTok{(sur.slm_lag11, }\DataTypeTok{tr =}\NormalTok{ trsur,}\DataTypeTok{R =} \DecValTok{1000}\NormalTok{)}

\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{1}\NormalTok{]]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Impact measures (lag, trace):
##                                       Direct   Indirect      Total
## log(GDPpc)_1                       0.6769737  0.6361932  1.3131669
## log(Older)_1                      -0.8042340 -0.7557874 -1.5600214
## log(Density)_1                     0.1670754  0.1570109  0.3240864
## Transit_1                          0.3430969  0.3224289  0.6655259
## log(Humidity_lag11)_1             -1.5656410 -1.4713276 -3.0369685
## log(Mean_Temp_lag11)_1            -1.7452380 -1.6401057 -3.3853437
## log(Sunshine_Hours_lag11 + 0.1)_1 -0.2818768 -0.2648967 -0.5467734
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{2}\NormalTok{]]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Impact measures (lag, trace):
##                                        Direct    Indirect      Total
## log(Density)_2                     0.06263494  0.04033489  0.1029698
## Transit_2                          0.44110065  0.28405466  0.7251553
## log(Humidity_lag11)_2             -1.06494878 -0.68579283 -1.7507416
## log(Mean_Temp_lag11)_2            -1.48108018 -0.95376810 -2.4348483
## log(Sunshine_Hours_lag11 + 0.1)_2 -0.19117818 -0.12311261 -0.3142908
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{impacts.sur.slm_lag11[[}\DecValTok{3}\NormalTok{]]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Impact measures (lag, trace):
##                                        Direct    Indirect       Total
## log(Density)_3                     0.02434857  0.01918949  0.04353807
## Transit_3                          0.43185937  0.34035519  0.77221456
## log(Humidity_lag11)_3             -0.97725083 -0.77018681 -1.74743764
## log(Mean_Temp_lag11)_3            -1.39946579 -1.10294110 -2.50240689
## log(Sunshine_Hours_lag11 + 0.1)_3 -0.09717791 -0.07658744 -0.17376535
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{myimpact <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DataTypeTok{ncol =} \DecValTok{5}\NormalTok{, }\DataTypeTok{nrow =} \DecValTok{29}\NormalTok{)}
\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{2}\OperatorTok{:}\DecValTok{30}\NormalTok{)\{}
\NormalTok{  myimpact[i}\DecValTok{-1}\NormalTok{,] <-}\StringTok{ }\NormalTok{impacts.sur.slm_lag11[[i]]}\OperatorTok{$}\NormalTok{res}\OperatorTok{$}\NormalTok{direct}
\NormalTok{\}}
\KeywordTok{matplot}\NormalTok{(myimpact, }\DataTypeTok{type=}\StringTok{'l'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/impacts-1.pdf}

@@

\leavevmode\hypertarget{ref-prem2017projecting}{}%
Prem, K., Cook, A.R., Jit, M., 2017. Projecting social contact matrices
in 152 countries using contact surveys and demographic data. PLoS
computational biology 13, e1005697.

@@

\leavevmode\hypertarget{ref-singh2020age}{}%
Singh, R., Adhikari, R., 2020. Age-structured impact of social
distancing on the covid-19 epidemic in india. arXiv preprint
arXiv:2003.12055.
